---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

Set working directory to correct path
```{r}
setwd("~/DataScience/Git/RepData_PeerAssessment1")
```

and load the data from the *activity.zip* file into an object *activityData*.
Convert dates using *as.Date()* function.

```{r}
activityData<-read.csv(unz("activity.zip", "activity.csv"))
activityData$date<-as.Date(activityData$date)
```

## What is mean total number of steps taken per day?

First we split the *activityData* dataset by day

```{r}
dayData<-split(activityData,activityData$date)
```

and now we compute the mean steps by each day and return it.

```{r}
daySteps<-sapply(dayData,function(x){sum(x$steps,na.rm=T)})
daySteps
```

Now we plot a histogram of the data in *daySteps*

```{r}
hist(daySteps,main="Histogram of Number of Steps per day",
     xlab="Number of Steps",freq=F,col="lightblue")
```

And finally we report the mean and median of the number of steps per day

```{r}
sapply(dayData,function(x){summary(x$steps,na.rm=T)[3:4]})
```

## What is the average daily activity pattern?

To answer that question first we are going to compute the average number of
steps taken for each 5-min interval over all days and then plot the resulting
time series

```{r}
intData<-split(activityData,activityData$interval)
intSteps<-sapply(intData,function(x){mean(x$steps,na.rm=T)})
plot(names(intSteps),intSteps,type="l",xlab="5-min interval",ylab="Mean steps",
     main="Time series of mean number of steps for 
     each 5-min time interval across all days")
```

We are also reporting the 5-min time interval with the highest mean number of
steps across all days.

```{r}
names(intSteps)[which(intSteps==max(intSteps))]
```

## Imputing missing values

We have noted that there are several values coded as *NA* indicating the 
presence of missing values in the *steps* variable. First we are going to
compute how many missing values do we have

```{r}
sum(is.na(activityData$steps))
```

We are going to fill all the missing values using the mean of the 5-min
interval in the new data frame *noNAData*.

```{r}
df<-data.frame(interval=names(intSteps),mean=intSteps)
merged<-merge(activityData,df,by="interval")
merged[is.na(merged$steps),]$steps<-merged[is.na(merged$steps),]$mean
ordered<-merged[order(merged$date,merged$interval,decreasing=F),]
noNAData<-data.frame(steps=ordered$steps,date=ordered$date,interval=ordered$interval)
```

Now with this new dataset we are going to compute the mean number of steps per
day and plot a histogram and report the mean and median too to compare with the
values obtained when there was missing data in the data set.

```{r}
dayData2<-split(noNAData,noNAData$date)
daySteps2<-sapply(dayData2,function(x){sum(x$steps)})
hist(daySteps2,main="Histogram of Number of Steps per day",
     xlab="Number of Steps",freq=F,col="lightblue")
sapply(dayData2,function(x){summary(x$steps)[3:4]})
```

We can see how imputting the NA values has made that some of the most extreme
values in the 0-5000 bin have moved to the central 10000-15000 bin.

## Are there differences in activity patterns between weekdays and weekends?

To study the differences between weekdays and weekends we are going to add a 
factor variable to out *noNAData* dataset to account for that difference.

```{r,message=FALSE}
library(lubridate)
noNAData$dayfactor<-as.factor(ifelse(wday(noNAData$date)==7 | 
                                 wday(noNAData$date)==1,"weekend","weekday"))
```

Now we will plot the average number of steps taken for each 5-min interval 
over all weekdays and weekends.

```{r}
dayfactorData<-split(noNAData,noNAData$dayfactor)
dayfactorwdayData<-dayfactorData$weekday
dayfactorwendData<-dayfactorData$weekend
splitwdayData<-split(dayfactorwdayData,dayfactorwdayData$interval)
splitwendData<-split(dayfactorwendData,dayfactorwendData$interval)
meanStepswday<-sapply(splitwdayData,function(x){mean(x$steps)})
meanStepswend<-sapply(splitwendData,function(x){mean(x$steps)})

min.width=trunc(min(as.numeric(names(meanStepswday)),
                    as.numeric(names(meanStepswend))))
max.width=ceiling(max(as.numeric(names(meanStepswday)),
                    as.numeric(names(meanStepswend))))
min.length=trunc(min(meanStepswday,meanStepswend))
max.length=ceiling(max(meanStepswday,meanStepswend))
par(mfrow=c(2,1))
par(cex = 0.6)
par(mar = c(0, 0, 0, 0), oma = c(4, 4, 0.5, 2))
par(tcl = -0.25)
par(mgp = c(2, 0.6, 0))
plot(names(meanStepswday),meanStepswday,type="l",xlab="",ylab="",axes=FALSE,
     xlim=c(min.width,max.width),ylim=c(min.length,max.length))
mtext("Weekdays",side=3,line=-1,adj = 0.1, cex = 0.6)
axis(4, at = seq(min.length, max.length, length.out=4))
box()
plot(names(meanStepswend),meanStepswend,type="l",xlab="",ylab="",axes=FALSE,
     xlim=c(min.width,max.width),ylim=c(min.length,max.length),)
mtext("Weekends",side=3,line=-1,adj = 0.1, cex = 0.6)
axis(2, at = seq(min.length, max.length, length.out=4))
axis(1, at = seq(min.width, max.width, length.out=4))
box()
mtext("5-min interval", side=1, outer=T, at=0.5,cex = 0.7, line = 2.2)
mtext("Mean steps", side=2, outer=T, at=0.5,cex = 0.7, line = 2.2)
```
